### 2018-03-14 submission/manuscript.Rmd
further edits to intro

This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies within the context of the overall field.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs]. Many of these sources of biases are made worse as cycle number increases [ref]. For example, the selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases [refs]. Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen [@Haas2011]. There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. With respect to primers, there are different binding affinities for different bacteria that are dependent on variable region [refs]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

>>>>

This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. First, it is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the borader context of the overall field.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [refs]. Many of these sources of biases are made worse as cycle number increases [ref]. For example, the selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [refs]. Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].

There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus Influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [refs]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases. There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study, we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-03-14 submission/manuscript.tex
further edits to intro

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that primers used,
cycle number, selective amplification, amplification error, non-specific
amplification, chimeras, and polymerase dependent differences can all
change the observed results {[}refs{]}. Many of these sources of biases
are made worse as cycle number increases {[}ref{]}. For example, the
selective amplification of the dominant sequence can increase the
difference between the most and next abundant sequence, over-emphasizing
the actual difference {[}ref{]}. Both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) can
also increase as cycle number increases {[}refs{]}. Additionally,
chimeras can form from an aborted extension followed by subsequent
priming error and secondary extension and even if it has not been
explicitly shown to increase with cycle number it is easy to postulate
that this does happen (Haas et al., 2011). There are also intrinsic
properties to primers and DNA polymerases chosen that can introduce
bias. With respect to primers, there are different binding affinities
for different bacteria that are dependent on variable region {[}refs{]}.
Additionally, there are multiple families of DNA polymerases that have
their own error rate and proof reading capcity (Ishino \& Ishino, 2014).
Interestingly, the influence that these different DNA polymerases can
have on the observed results have not been well studied like some of the
other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that optimization of the PCR protocol could
reduce error and chimera generation (Gohl et al., 2016). This study also
found that, regardless of DNA polymerase, the number of Operational
Taxonomic Units (OTUs) or taxa generated were not easily reduced using
the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
natural to extend this line of inquiry and ask if biases in the number
of OTUs and chimeras are also dependent on HiFi DNA polymerases. There
is some reason to think that this may be the case since many of these
HiFi DNA polymerase come from different families (e.g. \emph{Taq}
belongs to the family A polymerases) and may intrinsically have
different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014). In this study we critically
examine if any of five different HiFi DNA polymerases introduce
significant biases into 16S rRNA gene surveys, if this is a cycle
dependent phenomenon, and whether they can be removed using a standard
bioinformatic pipeline.

>>>>

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results
{[}refs{]}. Many of these sources of biases are made worse as cycle
number increases {[}ref{]}. For example, the selective amplification of
the dominant sequence can increase the difference between the most and
next abundant sequence, over-emphasizing the actual difference
{[}ref{]}. Both amplification error and non-specific amplification
(e.g.~incorrect amplicon size products) can also increase as cycle
number increases which can drastically change commonly used diversity
measures {[}refs{]}. Additionally, chimeras can form from an aborted
extension followed by subsequent priming error and secondary extension
and even if it has not been explicitly shown to increase with cycle
number it is easy to postulate that this does happen and will also
artificially increase community diversity (Haas et al., 2011).

There are also intrinsic properties to primers and DNA polymerases
chosen that can introduce bias. Primers have variable region dependent
binding affinities for different bacteria and depending on the primer
pair do not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus Influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) {[}refs{]}. Additionally, there are
multiple families of DNA polymerases that have their own error rate and
proof reading capcity (Ishino \& Ishino, 2014). Interestingly, the
influence that these different DNA polymerases can have on the observed
results have not been well studied like some of the other previously
mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that optimization of the PCR protocol could
reduce error and chimera generation (Gohl et al., 2016). This study also
found that regardless of DNA polymerase, the number of Operational
Taxonomic Units (OTUs) or taxa generated were not easily reduced using
the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
natural to extend this line of inquiry and ask if biases in the number
of OTUs and chimeras are also dependent on HiFi DNA polymerases. There
is some reason to think that this may be the case since many of these
HiFi DNA polymerases come from different families (e.g. \emph{Taq}
belongs to the family A polymerases) and may intrinsically have
different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014). In this study, we critically
examine if any of five different HiFi DNA polymerases introduce
significant biases into 16S rRNA gene surveys, if this is a cycle
dependent phenomenon, and whether they can be removed using a standard
bioinformatic pipeline.
